\section{Preliminaries and Problem Formulation}\label{sec:prelim}

\subsection{Kolmogorov--Arnold Network Layer}

Kolmogorov--Arnold Networks (KANs) \cite{liu2024kan} replace the fixed neuron-wise activations of an MLP with \emph{learnable} edge-wise activations parameterised as B-splines. Formally:

\begin{definition}[KAN Layer]\label{def:kan}
Given input $\mathbf{x} \in \mathbb{R}^{n_{\text{in}}}$, grid size $G \in \mathbb{N}$, and B-spline order $k \in \mathbb{N}$, a KAN layer is the map $\Phi: \mathbb{R}^{n_{\text{in}}} \to \mathbb{R}^{n_{\text{out}}}$
\begin{equation}
\Phi(\mathbf{x})_q \;=\; \sum_{p=1}^{n_{\text{in}}} \phi_{q,p}(x_p), \qquad q = 1, \ldots, n_{\text{out}},
\end{equation}
with each edge function
\begin{equation}\label{eq:edge}
\phi_{q,p}(x) \;=\; w^{(b)}_{q,p}\, \sigma(x) \;+\; w^{(s)}_{q,p}\, \sum_{i=1}^{G+k} c_{q,p,i}\, B_i(x),
\end{equation}
where $\sigma(\cdot) = \mathrm{SiLU}(\cdot)$ is the residual base activation, $\{B_i\}_{i=1}^{G+k}$ is a B-spline basis of order $k$ on a uniform grid of $G$ points, and $\Theta_{\text{KAN}} = \{w^{(b)}_{q,p}, w^{(s)}_{q,p}, c_{q,p,i}\}$ are learnable.
\end{definition}

By contrast, a fully-connected MLP layer with the same $(n_{\text{in}}, n_{\text{out}})$ employs a dense weight matrix $W \in \mathbb{R}^{n_{\text{out}} \times n_{\text{in}}}$, a bias $\mathbf{b} \in \mathbb{R}^{n_{\text{out}}}$, and a fixed scalar nonlinearity, so $\Theta_{\text{MLP}} = \{W, \mathbf{b}\}$. The architectural distinction relevant to our analysis is that each scalar weight in $W$ controls a global linear interaction, whereas each spline coefficient $c_{q,p,i}$ has \emph{local} support on the grid, by construction of the B-spline basis.

\subsection{Communication-Constrained Federated IDS}

\begin{definition}[Communication-Constrained Federated IDS]\label{def:fl-ids}
Consider $K$ IoT clients. Client $C_k$ holds a private dataset $\mathcal{D}_k = \{(\mathbf{x}_i, y_i)\}_{i=1}^{N_k}$ with feature vectors $\mathbf{x}_i \in \mathbb{R}^d$ and labels $y_i \in \{0, 1, \ldots, c-1\}$, drawn independently from a per-client distribution $\mathcal{P}_k$. Heterogeneity across clients is modelled by drawing the per-class proportions of $\mathcal{P}_k$ from a Dirichlet distribution with concentration $\alpha > 0$ \cite{hsu2019measuring}: small $\alpha$ yields highly skewed $\mathcal{P}_k$ (extreme non-IID), while $\alpha \to \infty$ recovers the IID setting.

Given a model $f(\Theta; \cdot)$ and a per-client empirical loss
\begin{equation}
\mathcal{L}_k(\Theta) \;=\; \frac{1}{N_k}\sum_{i=1}^{N_k} \ell\big(f(\Theta; \mathbf{x}_i),\, y_i\big),
\end{equation}
the federated IDS objective is
\begin{equation}\label{eq:fl-obj}
\min_{\Theta} \; F(\Theta) \;=\; \sum_{k=1}^{K} \frac{N_k}{N}\, \mathcal{L}_k(\Theta) \quad\text{s.t.}\quad \sum_{t=1}^{T} \mathcal{C}_t \;\leq\; \mathcal{B},
\end{equation}
where $N = \sum_k N_k$, $\mathcal{C}_t$ is the per-round bytes summed across participating clients, and $\mathcal{B}$ is the deployment bandwidth budget.
\end{definition}

In a synchronous FedAvg \cite{mcmahan2017fedavg} solver of \eqref{eq:fl-obj}, each round $t$ samples a fraction $C \in (0, 1]$ of the clients, transmits the global parameters $\Theta_t$ to each, runs $E$ local epochs of stochastic gradient descent, and aggregates the resulting local parameters $\{\Theta_{t+1}^k\}$ by data-weighted averaging.

\subsection{Related Work}\label{subsec:related}

We position FedKAN-IDS at the intersection of three active research threads.

\textbf{(i) FL for IoT intrusion detection.} DeepFed \cite{li2020deepfed} introduced a CNN-GRU architecture for federated industrial IDS, demonstrating that FL can match centralised detection rates while preserving data locality. Subsequent surveys \cite{nguyen2021fl-iot-survey} catalogued the design space of FL-IDS and identified two persistent bottlenecks: communication overhead on bandwidth-constrained gateways, and accuracy degradation under non-IID partitioning of network traffic. Robust FL methods \cite{sattler2019sparse} have addressed the latter by reducing the effective heterogeneity through compressed communication, but the underlying client architecture has remained, almost without exception, a dense MLP or CNN.

\textbf{(ii) Kolmogorov--Arnold Networks and their federation.} The KAN architecture of Liu \emph{et al.} \cite{liu2024kan} replaces fixed neuron-wise activations with learnable B-spline edge activations, providing better interpretability and parameter efficiency on smooth scientific-computing benchmarks. Within a year of that proposal, several works ported KAN to federated learning. F-KANs \cite{zeydan2024fkans} integrated KAN into vanilla FedAvg on CIFAR-style image benchmarks. Sasse and de Farias \cite{sasse2024evaluating} performed the first systematic non-IID evaluation of FedKAN on tabular data, reporting modest accuracy gains over MLP at matched parameter budgets and proposing the same parameter-matching protocol we adopt here. Ma \emph{et al.} \cite{ma2025enhancing} extended the comparison across multiple aggregation strategies (FedProx, SCAFFOLD), and Zeydan \emph{et al.} \cite{zeydan2025fedkan} applied a Fed-KAN to traffic prediction. Lee \emph{et al.} \cite{lee2025unified} contributed a medical-imaging benchmark and Zeleke and Bochicchio \cite{zeleke2024federated} an ECG-signal study. Most recently, Yu \emph{et al.} introduced CG-FKAN \cite{yu2025cgfkan}, which sparsifies KAN grid coefficients to attack communication efficiency directly.

The methodological observation underlying this body of work --- that FedKAN converges in fewer rounds and transmits fewer bytes than a parameter-heavy MLP --- has therefore been published several times outside the IDS setting. Our contribution is \emph{not} to claim that observation as new. Rather, we contribute (a) the first systematic study of FedKAN on the standardised NetFlow-v2 family of IDS datasets~\cite{sarhan2022standard}, (b) a parameter-matched experimental protocol that resolves an unfair-baseline ambiguity present in prior FedKAN work, and (c) a refined empirical claim --- that FedKAN's advantage is robustness/tail-risk under heterogeneity rather than mean accuracy --- anchored in a convergence analysis with an explicit spline-bias term.

\textbf{(iii) Communication-efficient FL.} Our framing also intersects with the broader literature on communication-efficient FL. FedPAQ \cite{reisizadeh2020fedpaq} reduces uplink bandwidth via periodic averaging plus stochastic quantisation; Top-$k$ SGD \cite{aji2017topk} sparsifies gradient updates to the largest $k$ entries; and \cite{sattler2019sparse} combined sparsification with non-IID--aware aggregation. CG-FKAN \cite{yu2025cgfkan} can be viewed as a KAN-specific instantiation of this line. Our protocol does \emph{not} apply any compression; we report all results in raw float32. As Section~\ref{sec:experiments} makes explicit, under parameter parity FedKAN is in fact slightly \emph{more} expensive to transmit than a comparable MLP, so the value proposition of our work is not communication efficiency per se but the heterogeneity-robust accuracy floor that FedKAN provides.
